Neighbor-Based Decentralized Training Strategies for Multi-Agent Reinforcement Learning

### 2024-10-12 paper-2024-sac-marl-neighbouring-strategies.tex
shrink algo description

In this paper, we focus on cooperative MARL scenarios, 
where agents must collaborate to achieve a common goal.
In the following, we provide an overview of the key concepts in MARL, 
including the formalization of the problem, the deep Q-learning algorithm, and the training paradigms used in the literature.
\subsection{Formalization}
Since MARL is a broad field, it is essential to provide a formalization to guide the discussion.
In this paper, we consider partially observable networked markov decision process (PONMDP)~\cite{DBLP:journals/tac/AdlakhaLG12} as a tuple $(\mathcal{G}, \mathcal{S}, \mathcal{A}, \mathcal{O}, \mathcal{P}, \mathcal{R}, \gamma)$, where

\begin{itemize}
  \item $\mathcal{G} = (N, E)$ is a communication graph, where $N$ is the set of $n$ agents and $E \subseteq N \times N$ represents the communication links between agents. 
  Time-varying graphs $\mathcal{G}_t = (N, E_t)$ can be used to represent communication evolving over time $t$.
  \item $\mathcal{S}$ is the global state space.
  \item $\mathcal{A} = \mathcal{A}^1 \times \dots \times \mathcal{A}^n$ is the joint action space, where $\mathcal{A}^i$ is the action space of agent $i$.
  \item $\mathcal{O} = \mathcal{O}^1 \times \dots \times \mathcal{O}^n$ is the joint observation space, where $\mathcal{O}^i$ is the observation space for agent $i$.
  \item $\mathcal{P}: \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is the state transition function, describing the probability of transitioning to a new state $s' \in \mathcal{S}$ given the current state $s \in \mathcal{S}$ and joint action $a \in \mathcal{A}$.
  \item $\mathcal{R} = {\mathcal{R}^i}_{i \in N}$, where $\mathcal{R}^i: \mathcal{S} \times \mathcal{A} \to \mathbb{R}$ is the reward function for agent $i$.
  \item $\gamma \in [0, 1]$ is the discount factor.
\end{itemize}
Each agent $i$ at time $t$ receives an observation $o^i_t \in \mathcal{O}^i$, 
takes an action $a^i_t \in \mathcal{A}^i$ based on its local policy $\pi^i: \mathcal{O}^i \times \mathcal{A}^i \to [0,1]$, and receives a reward $r^i_{t+1} = \mathcal{R}^i(s_t, a_t)$. 
%
The global state evolves according to $\mathcal{P}$. 
Agents can exchange information with their neighbors in $\mathcal{G}$ 
(or $\mathcal{G}t$ if the graph is time-varying). 
%
The objective is typically to maximize the expected discounted sum of rewards $\sum{t=0}^{\infty} \gamma^t r^i_t$ for each agent $i$, 
or some global reward function like the average reward over all agents.
%Formally, it can be defined as:
%\[ J(\{\pi^i\}) = \lim_{T \to \infty} \mathbb{E} \left[ \frac{1}{T} \sum_{t=0}^{T-1} \frac{1}{N} \sum_{i \in \mathcal{N}} R^i(s_t, a_t) \right] \]
Agents typically use other function to estimate the value of taking an action in a given state, such as the action-value function (also known as \emph{Q-value}). 
The latter represents the expected cumulative discounted reward the agent will receive by taking action $a \in \mathcal{A}$ in state $s \in \mathcal{S}$ and then following its policy $\pi^i$:

@@

\paragraph{Centralized Training with Decentralized Execution (CTDE)}
In CTDE, a central learner trains all agents, 
but execution of the learned policy is decentralized. 
This process typically involves two phases:
\begin{itemize}
  \item \textit{Offline training:} Agents interact with a simulator or environment, gathering experience data used to update the central learner's policy.
  \item \textit{Online execution:} Agents deploy the learned policy in the environment without further learning.
\end{itemize}
This approach offers several benefits.  
Leveraging a centralized dataset allows for efficient learning, 
potentially exploiting global information, 
which is crucial in cooperative scenarios requiring coordinated actions. 
Furthermore, decentralized execution allows agents to act independently after training.
%
However, CTDE has limitations. 
Scalability can become problematic with a large number of agents. 
The reliance on a central learner hinders adaptation to dynamic environments or tasks during runtime.
\sloppy
\paragraph{Decentralized Training with Decentralized Execution (DTDE)}
DTDE eliminates the central learner; each agent learns independently from its own experiences.
Inter-agent cooperation and communication can be incorporated during both training and execution. 
Unlike CTDE, training and execution phases are not necessarily separate---simultaneous learning and execution allow for adaptation to 
changing environments or tasks at runtime.
%
Key advantages of DTDE include scalability with the number of agents and adaptability to dynamic environments.
%
However, DTDE also presents several challenges to overcome, like: 
i) convergence time is often longer compared to centralized training. 
ii) The lack of central coordination can lead to instability and difficulties in differentiating agents and their environmental interactions; and 
iii) online learning in DTDE can introduce non-stationarity if agents do not account for other agents' evolving policies during training, 
as the environment effectively changes from each agent's perspective.

@@

In \Cref{alg:neighbor_dqn} we provide a detailed description of the integration of the neighboring-based strategies with deep Q-learning.
\begin{algorithm}
  \caption{Neighboring-Based Deep Q-Learning}
  \label{alg:neighbor_dqn}
  \ForEach{agent $i \in N$}{
      Initialize Q-network $Q^i(o^i, a^i; \theta^i)$ and target network $Q^{i-}(o^i, a^i; \theta^{i-})$\\
      Initialize replay buffer $\mathcal{D}^i$
  }
  \For{episode = 1 to M}{
      \ForEach{agent $i \in N$}{
          Initialize initial observation $o_0^i$
      }
      \For{t = 0 to T}{
          \ForEach{agent $i \in N$}{
              Select action $a_t^i$ based on $\epsilon$-greedy policy using $Q^i(o_t^i, a^i; \theta^i)$\\
              Execute action $a_t^i$ and observe reward $r_{t+1}^i$ and next observation $o_{t+1}^i$\\
              Store $(o_t^i, a_t^i, r_{t+1}^i, o_{t+1}^i)$ in $\mathcal{D}^i$\\
              Sample a mini-batch from $\mathcal{D}^i$\\
              Calculate target values using chosen neighboring strategy (see below)\\
              Update $\theta^i$ by minimizing the TD error\\
              Every C steps, update target network $\theta^{i-} \leftarrow \theta^i$\\
              \textbf{
              Every S steps, perform neighbor interaction (e.g., averaging, consensus, experience sharing)}
          }
      }
  }
\end{algorithm}
\section{Evaluation}\label{sec:experiments}
This section presents the experimental setup and evaluation metrics used to assess the performance of the proposed neighboring-based training strategies.
All the experiments and data produced are reproducible and publicly available\footnote{\url{https://anonymous.4open.science/r/experiments-2024-SAC-marl-neighbouring-strategies-4B8F}}.
\subsection{Simulated environment}
The proposed multi-agent environment is a variant of the ``gather the items'' scenario, designed to study the impact of neighborhood strategies. 
It consists of a 2D Euclidean space populated by $A$ agents and $T$ targets, 
initially placed at random locations.
%
The objective is for the agents to collectively capture all targets, 
at which point the environment episode terminates successfully.  
A maximum time limit of $T_{max} = 300$ time steps is enforced to prevent indefinite episodes.

@@

\begin{figure}
  \centering
  \includegraphics[width=1\linewidth]{figures/behavior.pdf}
  \caption{Examples of cooperative and uncoordinated behavior. (Left) desired, agents effectively choose different paths to collect different items. (Right) undesired, agents cover the same path to collect the same item.}
  \label{fig:behavior}
\end{figure}

>>>>

%In this paper, we focus on cooperative MARL scenarios, 
%where agents must collaborate to achieve a common goal.
%In the following, we provide an overview of the key concepts in MARL, 
%including the formalization of the problem, the deep Q-learning algorithm, and the training paradigms used in the literature.
\subsection{Formalization}
Since MARL is a broad field, it is essential to provide a formalization to guide the discussion.
In this paper, we consider partially observable networked markov decision process (PONMDP)~\cite{DBLP:journals/tac/AdlakhaLG12} as a tuple $(\mathcal{G}, \mathcal{S}, \mathcal{A}, \mathcal{O}, \mathcal{P}, \mathcal{R}, \gamma)$, where

\begin{itemize}
  \item $\mathcal{G} = (N, E)$ is a communication graph, where $N$ is the set of $n$ agents and $E \subseteq N \times N$ represents the communication links between agents. 
  Time-varying graphs $\mathcal{G}_t = (N, E_t)$ can be used to represent communication evolving over time $t$.
  \item $\mathcal{S}$ is the global state space.
  \item $\mathcal{A} = \mathcal{A}^1 \times \dots \times \mathcal{A}^n$ is the joint action space, where $\mathcal{A}^i$ is the action space of agent $i$.
  \item $\mathcal{O} = \mathcal{O}^1 \times \dots \times \mathcal{O}^n$ is the joint observation space, where $\mathcal{O}^i$ is the observation space for agent $i$.
  \item $\mathcal{P}: \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is the state transition function, describing the probability of transitioning to a new state $s' \in \mathcal{S}$ given the current state $s \in \mathcal{S}$ and joint action $a \in \mathcal{A}$.
  \item $\mathcal{R} = {\mathcal{R}^i}_{i \in N}$, where $\mathcal{R}^i: \mathcal{S} \times \mathcal{A} \to \mathbb{R}$ is the reward function for agent $i$.
  \item $\gamma \in [0, 1]$ is the discount factor.
\end{itemize}
Each agent $i$ at time $t$ receives an observation $o^i_t \in \mathcal{O}^i$, 
takes an action $a^i_t \in \mathcal{A}^i$ based on its local policy $\pi^i: \mathcal{O}^i \times \mathcal{A}^i \to [0,1]$, and receives a reward $r^i_{t+1} = \mathcal{R}^i(s_t, a_t)$. 
%
The global state evolves according to $\mathcal{P}$. 
Agents can exchange information with their neighbors in $\mathcal{G}$ 
(or $\mathcal{G}t$ if the graph is time-varying). 
%
The objective is typically to maximize the expected discounted sum of rewards $\sum{t=0}^{\infty} \gamma^t r^i_t$ for each agent $i$, 
or some global reward function like the average reward over all agents in case of cooperative scenarios.
%Formally, it can be defined as:
%\[ J(\{\pi^i\}) = \lim_{T \to \infty} \mathbb{E} \left[ \frac{1}{T} \sum_{t=0}^{T-1} \frac{1}{N} \sum_{i \in \mathcal{N}} R^i(s_t, a_t) \right] \]
Agents typically use other function to estimate the value of taking an action in a given state, such as the action-value function (also known as \emph{Q-value}). 
The latter represents the expected cumulative discounted reward the agent will receive by taking action $a \in \mathcal{A}$ in state $s \in \mathcal{S}$ and then following its policy $\pi^i$:

@@

\paragraph{Centralized Training with Decentralized Execution (CTDE)}
In CTDE, a central learner trains all agents, 
but execution of the learned policy is decentralized. 
This process typically involves two phases:
\begin{enumerate}
  \item \textit{Offline training:} Agents interact with a simulator or environment, gathering experience data used to update the central learner's policy.
  \item \textit{Online execution:} Agents deploy the learned policy in the environment without further learning.
\end{enumerate}
This approach offers several benefits.  
Leveraging a centralized dataset allows for efficient learning, 
potentially exploiting global information, 
which is crucial in cooperative scenarios requiring coordinated actions. 
Furthermore, decentralized execution allows agents to act independently after training.
%
However, CTDE has limitations. 
Scalability can become problematic with a large number of agents. 
The reliance on a central learner hinders adaptation to dynamic environments or tasks during runtime.
\sloppy
\paragraph{Decentralized Training with Decentralized Execution (DTDE)}
DTDE eliminates the central learner; each agent learns independently from its own experiences.
Inter-agent cooperation and communication can be incorporated during both training and execution. 
Unlike CTDE, training and execution phases are not necessarily separate---simultaneous learning and execution allow for adaptation to 
changing environments or tasks at runtime.
%
Key advantages of DTDE include scalability with the number of agents and adaptability to dynamic environments.
%
However, DTDE also presents several challenges to overcome, like: 
i) convergence time is often longer compared to centralized training. 
ii) The lack of central coordination can lead to instability and difficulties in differentiating agents and their environmental interactions; and 
iii) online learning in DTDE can introduce non-stationarity if agents do not account for other agents' evolving policies during training, 
as the environment effectively changes from each agent's perspective.

@@

In \Cref{alg:neighbor_dqn} we provide a detailed description of the integration of the neighboring-based strategies with deep Q-learning.
\begin{algorithm}
  \caption{Neighboring-Based Deep Q-Learning}
  \label{alg:neighbor_dqn}
  \ForEach{agent $i \in N$}{
      Initialize Q-network $Q^i(o^i, a^i; \theta^i)$ and target network $Q^{i-}(o^i, a^i; \theta^{i-})$\\
      Initialize replay buffer $\mathcal{D}^i$
  }
  \For{episode = 1 to M}{
      \ForEach{agent $i \in N$}{
          Initialize initial observation $o_0^i$
      }
      \For{t = 0 to T}{
          \ForEach{agent $i \in N$}{
              Select action $a_t^i$ based on $\epsilon$-greedy policy using $Q^i(o_t^i, a^i; \theta^i)$\\
              Execute action $a_t^i$ and observe reward $r_{t+1}^i$ and next observation $o_{t+1}^i$\\
              Store $(o_t^i, a_t^i, r_{t+1}^i, o_{t+1}^i)$ in $\mathcal{D}^i$\\
              Sample a mini-batch from $\mathcal{D}^i$\\
              Calculate target values using chosen neighboring strategy (see below)\\
              Update $\theta^i$ by minimizing the TD error\\
              Every C steps, update target network $\theta^{i-} \leftarrow \theta^i$\\
              \textbf{
              Every S steps, perform neighboring strategy}
          }
      }
  }
\end{algorithm}
\section{Evaluation}\label{sec:experiments}
This section presents the experimental setup and evaluation metrics used to assess the performance of the proposed neighboring-based training strategies.
All the experiments and data produced are reproducible and publicly available\footnote{\url{https://anonymous.4open.science/r/experiments-2024-SAC-marl-neighbouring-strategies-4B8F}}.
\subsection{Simulated environment}
The proposed multi-agent environment is a variant of the ``gather the items'' scenario, designed to study the impact of neighborhood strategies. 
It consists of a 2D Euclidean space populated by $A$ agents and $T$ targets, 
initially placed at random locations.
%
The objective is for the agents to collectively capture all targets, 
at which point the environment episode terminates successfully.  
A maximum time limit of $T_{max} = 300$ time steps is enforced to prevent indefinite episodes.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.7\linewidth]{figures/behavior.pdf}
  \caption{Examples of cooperative and uncoordinated behavior. (Left) desired, agents effectively choose different paths to collect different items. (Right) undesired, agents cover the same path to collect the same item.}
  \label{fig:behavior}
\end{figure}
